---
title: "Plotting and visualising data"
author:
- name: Laurent Gatto
  affiliation: de Duve Institute, UCLouvain
  email: laurent.gatto@uclouvain.be
output:
  BiocStyle::html_document
abstract: |
  Description of your vignette
---

# Introduction

This material is based on chapter 3 of the [*Modern Statistics for
Modern Biology*](https://www.huber.embl.de/msmb/) book by Susan Holmes
and Wolfgang Huber.

We want to visualise data to 

- interactively explore and understant data, i.e. *exploratory data
  analysis*;
- to present and *communicate results*, whether as a preliminary
  analysis or final results.

Our learning objectives are

- Familiarise ourselves with plotting in R using base graphics and
  `ggplot2`;
- Understand the concepts behind the gramar of graphics and `ggplot2`;
- Learn how to visualiste data along 1, 2 or more dimensions;
- Learn about `plotly`, one package for ploducing interactive
  visualisations.

# Base plotting

The default graphics system that comes with R, often called base R
graphics is simple and fast. It is based on the painter's model or
canvas, where different output are directly overlaid on top of each
other.

Below, we display the relation between the optical density of the
deoxyribonuclease (DNase) protein as measure by an enzyme-linked
immunosorbent assay (ELISA) assay for all observations.

```{r dnase, fig.cap = "The default base `plot` function on the `DNase` data."}
head(DNase)
plot(DNase$con, DNase$density)
```

We can add some features on the plot, such vertical dotted lines for
all observed observations and customise the look and feel of the plot
by setting specific arguments to the `plot` function.

```{r dnase2, fig.cap = "Customising a base figure using function arguments and overlaying new graphical features."}
plot(DNase$con, DNase$density,
     xlab = "DNase concentration (ng/ml)",
     ylab = "Optical density",
     pch = 1,
     col = "steelblue")
abline(v = unique(DNase$conc), lty = "dotted")
```

If we wanted to change anything to that figures, we would need to
repeat all the commands and modify accordingly. Any additinal command
would be added to the existing canvas.

Below, we use the `hist` and `boxplot` functions to produce and
histogram of all optical densities and a boxplot of the densities
split by run.

```{r hist_boxplot, fig.width = 14, fig.height = 7, fig.cap = "The base `hist` and `boxplot` functions, the latter using the formula interface."}
par(mfrow = c(1, 2))
hist(DNase$density)
boxplot(density ~ Run, DNase)
```

The base graphics function are very effective to quickly produce out
of the box figures. However, there is not global overview and
parametrisation of the visualisation. The layout decisions have to be
made up upfront (and if not adequate, the figure needs to be redrawn)
and every aspectr of the figure is customised locally as function
arguments.

More generally, base graphics functions will work with various inputs:
above we have wored with a `data.frame`, vectors and a formula. There
is no unified type of data across all functions which makes it
efficient for some types of data (if they match), but also very
heterogeneous in terms of interface, leading to a lot of customisation
code.

Finally, defaults, and colours in particular, are poorly chosen.

# `ggplot2`

**`ggplot2`** is a plotting package that makes it simple to create
complex plots from data in a data frame. It provides a more
programmatic interface for specifying what variables to plot, how they
are displayed, and general visual properties. The theoretical
foundation that supports the `ggplot2` is the *Grammar of
Graphics*[^gg]. Instead of producing the figure, the user defines and
assembles the visual components into an object that is the
displayed. There is a book about ggplot2[^ggbook] that provides a good
overview, but it is outdated. The ggplot2 webpage
(https://ggplot2.tidyverse.org) provides ample documentation.


[^gg]: Wilkinson, Leland. 2005. The Grammar of Graphics (Statistics
    and Computing). Berlin, Heidelberg: Springer-Verlag.

[^ggbook]: Wickham, Hadley. 2016. ggplot2: Elegant Graphics for Data
    Analysis. Springer-Verlag New York.


To build a ggplot, we will use the following basic template that can
be used for different types of plots:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

We need first to load the `ggplot2` package:

```{r ggload}
library("ggplot2")
```

- use the `ggplot()` function and bind the plot to a specific data frame using the
      `data` argument

```{r gg0, fig.cap = "We have only specified the data, and there's nothing to display yet."}
ggplot(data = DNase)
```

- define a mapping (using the aesthetic (`aes`) function), by
  selecting the variables to be plotted and specifying how to present
  them in the graph, e.g. as x/y positions or characteristics such as
  size, shape, color, etc.

```{r gg1, fig.cap = "`ggplot2` can now generate the axes, ticks and ranges based on the data."}
ggplot(data = DNase,
       mapping = aes(x = conc, y = density))
```

- add *geoms* – graphical representations of the data in the plot
  (points, lines, bars). To add a geom to the plot use the `+`
  operator. Because we have two continuous variables, let's use
  `geom_point()` first:

```{r gg2, fig.cap = "Final figures with rendering of the data as a scatter plot."}
ggplot(data = DNase,
       mapping = aes(x = conc, y = density)) + 
  geom_point()
```

It is possible to store the output of the `ggplot` function into a
variable that can be visualised by either typing its name in the
console or explicitly printing it (like any other variable).

```{r ggobj}
gg <- ggplot(data = DNase,
             mapping = aes(x = conc, y = density)) + 
    geom_point()
print(gg)
```

Let's immediately customise this visualisaion to 
- highlight how to re-use the `gg` object without repeating the
  plotting code and
- how we can add additional (identical or different) *goems* to a
  plot.

```{r gg3, fig.cap = "Adding another `geom_point` with its own (local) aesthetics. "}
gg + geom_point(aes(colour = Run))
```


# Visualising 1D data

# Visualising 2D data

# Visualising data along more dimensions

# Interactive visualisation

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
