---
title: 'Visualisation of high-dimensional data'
author: "Laurent Gatto"
output:
  rmdshower::shower_presentation:
    theme: material
    self_contained: true
    ratio: 16x10
---

## Visualisation of high-dimensional data 

```
Laurent Gatto                      Computational Proteomics Unit
https://lgatto.github.io           University of Cambridge
lg390@cam.ac.uk                    @lgatt0
```

Source: https://github.com/lgatto/visualisation


```{r env, warning=FALSE, echo=FALSE}
suppressPackageStartupMessages(library("pRoloc"))
suppressPackageStartupMessages(library("pRolocdata"))
suppressPackageStartupMessages(library("Rtsne"))
data(mulvey2015)
data(hyperLOPIT2015)
```

## Proteomics examples

Sample-level visualisations from
[Mulvey *et al.* (2015)](https://www.ncbi.nlm.nih.gov/pubmed/26059426)
*Dynamic Proteomic Profiling of Extra-Embryonic Endoderm
Differentiation in Mouse Embryonic Stem Cells.*

Protein-level visualisations from
[Christoforou *et al.* (2016)](https://www.ncbi.nlm.nih.gov/pubmed/26754106)
*A draft map of the mouse pluripotent stem cell spatial proteome.*

## Hierarchical clustering

Hierarchincal clustering start by calculating all pairwise distances
between all features (or samples) and then clusters/groups these based
on these similarities. There are various distances measures and
clustering algorithms that can be used. 

```{r}

```
## Principal Component Analysis

Principal Component Analysis (PCA) is a technique that transforms the
original n-dimentional data into a new n-dimentional space. Along
these new dimensions, called principal components, the data expresses
most of its variability along the first PC, then second, .... These
new dimensions are *linear combinations* of the orignal data.

Figures produces with `plot2D` function from the
[`pRoloc` package](https://bioconductor.org/packages/devel/bioc/html/pRoloc.html)

## {.fullpage}

```{r pcaex, echo=FALSE, fig.width = 12, fig.height = 4}
set.seed(1)
xy <- data.frame(x = (x <- rnorm(50, 2, 1)),
                 y = x + rnorm(50, 1, 0.5))
pca <- prcomp(xy)

z <- cbind(x = c(-1, 1), y = c(0, 0))
zhat <- z %*% t(pca$rotation[, 1:2])
zhat <- scale(zhat, center = colMeans(xy), scale = FALSE)
par(mfrow = c(1, 3))
plot(xy, main = "Orignal data (2 dimensions)")
plot(xy, main = "Orignal data with PC1")
abline(lm(y ~ x, data = data.frame(zhat - 10)), lty = "dashed")
grid()
plot(pca$x, main = "Data in PCA space")
grid()
```

## { .fullpage }

```{r pcaprot, echo=FALSE, fig.width=14, fig.height=7}
par(mfrow = c(1, 2))
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
.pca <- plot2D(t(mulvey2015), fcol = "times", cex = 3,
               sub = "Mulvey et al. (2015)",
               main = "Stem cell differentiation")
text(.pca[, 1], .pca[, 2], mulvey2015$rep)
legend("topleft", bty = "n", 
       legend = c(paste("Time", unique(mulvey2015$times)), "Replicates (1,2,3)"),
       pch = 19, col = c(getStockcol()[1:6], NA))
plot2D(hyperLOPIT2015,
       main = "Sub-cellular localisation map",
       sub = "Christoforou et al. (2016)")
addLegend(hyperLOPIT2015, cex = .8)
```


## t-Distributed Stochastic Neighbour Embedding

t-Distributed Stochastic Neighbour Embedding (t-SNE) is a *non-linear*
dimensionality reduction techique, i.e. that different regions of the
data space will be subjected to different transformations. In t-SNE
will compress small distances, thus bringing close neighbours
together, and will ignore large distances.

Figures produces with `plot2D` function from the
[`pRoloc` package](https://bioconductor.org/packages/devel/bioc/html/pRoloc.html)


## Parameter tuning

**Beware** of the `perplexity` and `steps` parameters, that should be
tuned before applying t-SNE.


## { .fullpage }

```{r tsneex, cache=TRUE, echo=FALSE, fig.width=14, fig.height=7}
par(mfrow = c(1, 2))
plot2D(hyperLOPIT2015, method = "PCA",
       sub = "Mulvey et al. (2015)",
       main = "Stem cell differentiation (PCA)")
plot2D(hyperLOPIT2015, method = "t-SNE",
       sub = "Mulvey et al. (2015)",
       main = "Stem cell differentiation (t-SNE)")
```

## References

### t-SNE
- https://lvdmaaten.github.io/tsne/
- http://distill.pub/2016/misread-tsne/

## License

This material is made available under the
[Creative Commons Attribution license](https://creativecommons.org/licenses/by/4.0/).

You are free to **share** - copy and redistribute the material in any
medium or format, and **adapt** - remix, transform, and build upon the
material for any purpose, even commercially it.
**Attribution** You must give appropriate credit, provide a link to
the license, and indicate if changes were made. You may do so in any
reasonable manner, but not in any way that suggests the licensor
endorses you or your use.

